---
title: "BDM Examples"
author: "Charles T T Edwards (NIWA, Wellington, New Zealand)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{bdm-examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.path='fig/bdm-examples-', fig.width=6, tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=70), message=FALSE, warning=FALSE, collapse = TRUE, comment = "#>")
```

# Example applications of the `bdm` R-package

```{r, results='hide'}
library(bdm)
```

## Chatham Rise Hake

The model is fitted to data from the chatham rise hake fishery in New Zealand, which consists of catches, a commerical abundance index and a survey index. The data are used to initialise an empirical data object (`edat`) that, by default, renormalises the indices to a geometric mean of one. See `?edat` for details.


```{r hakcr-data, results='hide', fig.cap='Chatham rise hake data'}
data(hakcr)
dat <- edat(harvest=hakcr$catch,index=cbind(hakcr$survey,hakcr$cpue),time=rownames(hakcr))
plot(dat)
```

### Development of a prior for $r$
Life history data are available, allowing us to populate an object of the `rdat` class. Monte-carlo samples are generated, and application of the `rcalc` function to this class of object calcuates values of $r$ for each iteration, producing an object of the `rprior` class. Finally, there is a `fit` function used to parameterise a log-normal distribution, and thereby to create an informative prior for the intrinsic growth rate. The estimated parameter values can be found in `object@lognormal.par`.
