---
title: "Example applications of `bdm`"
author: "Charles T T Edwards (NIWA, Wellington, New Zealand)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{bdm-examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.path='fig/bdm-examples-', fig.width=6, tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=75), message=FALSE, warning=FALSE, collapse = TRUE, comment = "#>")
```

Example applications of the `bdm` R-package are given here, based on data from fisheries in New Zealand.

```{r, results='hide'}
library(bdm)
```

# Chatham Rise Hake

The model is fitted to data from the chatham rise hake fishery in New Zealand, which consists of catches, a commerical abundance index and a survey index. The data are used to initialise an empirical data object (`edat`) that, by default, renormalises the indices to a arithmetic mean of one. See `?edat` for details.

```{r hakcr-data, results='hide', fig.cap='Chatham rise hake data'}
data(hakcr)
dat <- edat(harvest = hakcr$catch,index = cbind(hakcr$survey, hakcr$cpue), time = rownames(hakcr), renormalise = TRUE)
plot(dat)
```

## Development of a prior for $r$
Life history data are available, allowing us to populate an object of the `rdat` class. Monte-carlo samples are generated, and application of the `rcalc` function to this class of object calcuates values of $r$ for each iteration, producing an object of the `rprior` class. Finally, there is a `fit` function used to parameterise a log-normal distribution, and thereby to create an informative prior for the intrinsic growth rate. The estimated parameter values can be found in `object@lognormal.par`.

```{r hakrprior, results='hide', fig.cap='Prior for $r$ generated from life-history data for chatham rise hake.'}
# initialise lh data object for calculation of r with uncertainty
lhdat <- rdat(amax = 100, iter = 200)

# then life-history vectors can be assigned to each iteration
# with or without uncertainty
nmort(lhdat)    <- list(mu=0.18)
maturity(lhdat) <- c(0.0,0.01,0.02,0.06,0.14,0.28,0.50,0.72,0.86,0.94,0.98,0.99,1.00)
size(lhdat)     <- list(mu=list(Linf=106.5,k=0.229,t0=0.01))
mass(lhdat)     <- list(mu=list(a=1.88e-9,b=3.305))
sr(lhdat)       <- list(type='BH',mu=0.90,cv=0.10)

# calculate r prior and fit log-normal distribution
r <- rcalc(lhdat)
plot(r)
```

## Estimation of depletion
The model is initialised using the `bdm` command, with no arguments. Arguments can be supplied in the form or alternative model code, but by default initialisation loads a generalised surplus production model, which will be used for this example. Following initialisation of the model, the prior information on $r$, contained in the `rprior` object, can be loaded using the `update_prior` function, which takes as input the `bdm` and `rprior` objects. This function updates the model code directly, following which it must be compiled.

```{r, results='hide'}
mdl <- bdm()
mdl <- update_prior(mdl,r,compile=TRUE)
```

The default generalised surplus production model allows the depletion at maxium sustainable yield (MSY) to be fixed by the user. The depletion at MSY ($\phi$) forms part of the model inputs encoded in the `edat` object and can be accessed or assigned using the `shape` function. For chatham rise hake we assume that MSY occurs at 40\% of the carrying capacity (i.e. $\phi=0.4$). 

```{r} 
shape(dat) <- 0.4
```

The model can then be run using the `fit` function, which makes use of the R-package `rstan` to implement an MCMC fit.

```{r, results='hide'}
mdl <- fit(mdl,dat)
```

Trace, posterior histogram and cumulative density plots can be produced using the `traceplot`, `histplot` and `cumsumplot` functions.

```{r haktrace, fig.cap='Traceplots for chatham rise hake fit.'}
traceplot(mdl)
```
```{r hakhist, fig.cap='Posterior histograms for chatham rise hake fit.'}
histplot(mdl,par = c('r','logK','q'))
```
```{r hakcumsum, fig.cap='Cumulative posterior denisty plot for chatham rise hake fit.'}
cumsumplot(mdl,par = c('r','logK'))
```

## Model outputs

We can plot model outputs using the `dynplot` function, which by default plots the `depletion` trajectory over time. Currently it can also be used to visualise `biomass`, `surplus_production`, and the `harvest_rate`.

```{r hakdep, fig.cap='Estimated depletion for chatham rise hake.'}
dynplot(mdl)
```

Plotting functions `traceplot`, `histplot` and `dynplot` return graphical objects from the `ggplot2` package and can be modified before they are printed. For example, to add a title to the plot of surplus production and change the y-axis one could type:

```{r, fig.show='hide'}
gg <- dynplot(mdl, pars = 'surplus_production')
gg <- gg + ggtitle('Surplus Production')
print(gg)
```

We can use the function `refpoints` to extract the reference point information from the fitted model object as a `list` that contains `msy`, `depletion_at_msy` (which is equal to $\phi$), `biomass_at_msy` and `harvest_rate_at_msy`. These reference points are functions of $r$, $K$ and $\phi$, and therefore (with the exception of $\phi$ itself) have a posterior distribution. We can extract the median values and write them in a table using `pander::pandoc.table`: 

```{r hakrefpts, results = "asis"}
pander::pandoc.table(data.frame(lapply(refpoints(mdl),median)))
```

Similarly, the function `status` can be used to access posterior distributions of `current_biomass`, `current_depletion` and `current_harvest_rate`, which correspond to the final assessment year, i.e. `dat$time[dat$T]`. We could easily plot these as histograms, using for example:

```{r, fig.cap='Current status estimates for chatham rise hake.'}
sta <- status(mdl)
dfr <- reshape2::melt(sta)
gg <- ggplot(dfr) + 
    geom_histogram(aes(x=value)) + 
    facet_wrap(~L1, scales = 'free_x')
print(gg)
```
Note that the depletion is given relative to $K$. This is a more robust estimate of status than status relative to $MSY$. Nevertheless we could extract the status relative to $MSY$ based reference points by using a combination of the `refpoints()` and `status()` functions.

It is possible to produce kobe plots by using the `as.kobe` function, which creates a `data.frame` in the appropriate format for the `kobe` package. 

```{r}
assmt <- as.kobe(mdl)
```

The Kobe risk assessment framework has been developed by scientists at the International Commission for the Conservation of Atlantic Tunas, as a means of presenting status advice. The stock is assessed relative to $B_{MSY}$ and $H_{MSY}$. If $B < B_{MSY}$ then it is classified as overfished. If $H > H_{MSY}$ then overfishing is taking place. Status can be visualised as a kobe phase plot, which is produced using the `kobe::kobePhase` function.

```{r kobeplot, results="hide",fig.cap='Kobe plot of current stock status.'}
library(kobe)
kobePhase(subset(assmt,year==max(assmt$year))) +
  geom_point(aes(stock,harvest), alpha=0.1)
```

The `as.kobe` function can also be used to produce a summary of stock status over time; specifically the probability that the stock is in the green, red and yellow quadrants of the phase plane.

```{r kobesummary, results = "asis"}
assmt <- as.kobe(mdl, what = 'smry')
pander::pandoc.table(assmt)
```


# Chatham Rise Orange Roughy

```{r orhcr-data, results='hide', fig.cap='Chatham rise orange roughy data'}
data(orhcr)
dat <- edat(harvest=orhcr$catch,index=as.matrix(orhcr[,-1]),time=rownames(orhcr),renormalise = TRUE)
plot(dat)
```


```{r orhrprior, results='hide', fig.cap='Prior for $r$ generated from life-history data for chatham rise orange roughy.'}
# initialise lh data object for calculation of r with uncertainty
lhdat <- rdat(amax=130,iter=200)

# then life-history vectors can be assigned to each iteration
# with or without uncertainty
nmort(lhdat)    <- list(mu=0.045,cv=0.2)
maturity(lhdat) <- list(mu=list(acrit=35.67,delta=1.55))
size(lhdat)     <- list(mu=list(Linf=37.78,k=0.059,t0=-0.491))
mass(lhdat)     <- list(mu=list(a=8.0e-2,b=2.75))
sr(lhdat)       <- list(type='BH',mu=0.75,cv=0.10)

# calculate r prior and fit log-normal distribution
r <- rcalc(lhdat)
plot(r)
```

```{r, results='hide'}
mdl <- bdm()
mdl <- update_prior(mdl,r,compile = TRUE)
```

```{r, results='hide'}
mdl <- fit(mdl,dat)
```

Trace and posterior histogram plots produced using the `traceplot` and `histplot` functions.

```{r orhtrace, fig.cap='Traceplots for chatham rise orange roughy fit.'}
traceplot(mdl,par = c('r','logK'))
```
```{r orhhist, fig.cap='Posterior histograms for chatham rise orange roughy fit.'}
histplot(mdl,par = c('r','logK'))
```
The status estimates of the stock can be obtained in a similar manner to the reference points above.

```{r orhstatus, results = "asis"}
pander::pandoc.table(data.frame(lapply(status(mdl),median)))
```

Based on these we can perform projections exploring alternative constant harvest rate options. For example:

```{r orhproj}
mdl.project <- project(mdl, harvest = c(0.01, 0.025, 0.05), time = 100)
```


